---
title: "Gavrila_TSH_analysis"
author: "Henrique"
date: "2024-02-06"
output: html_document
---

```{r setup}
# BiocManager::install("DESeq2")
# # Load necessary libraries
library(DESeq2)
library(openxlsx)
library(ggpubr)
library(tidyverse)

#Solve conflicts
conflicted::conflict_prefer("rename", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")

# Formatting the metadata
meta <- read.xlsx("../data/Metadata_for_correlation.xlsx")

meta <- meta %>% separate_wider_delim(cols = Sample, delim = " ", names = c("Tissue", "Donor_ID"))
meta <- meta %>%
  rename(
    "Serum_Free_T3" = "Free.T3",
    "Serum_Free_T4" = "Free.T4",
    "Serum_TSH" = "TSH",
    "log10_Serum_TSH" = "log[TSH]"
  ) %>%
  mutate("log10_Serum_TSH" = log10(Serum_TSH))
names(meta) <- gsub("[ /]", "_", names(meta))

```

#Running updated DESEq2
## --- Create dds object ---
```{r}
# Running DESeq2
cts <- readRDS(file = "../data/tx2gene_counts.rds")
cts <- round(cts)
str(cts)

# Create a sample information table (coldata)
samples <- colnames(cts)
condition <- ifelse(grepl("^Deep", samples), "Deep", "SC") # Assign condition based on sample names
coldata <- data.frame(
  row.names = samples,
  condition = factor(condition, levels = c("SC", "Deep")) # Specifying factor levels
)

# Verify coldata matches cts
if (!all(rownames(coldata) == colnames(cts))) {
  stop("Row names of coldata must match column names of counts matrix")
}

# Create DESeqDataSet
dds <- DESeqDataSetFromMatrix(
  countData = cts,
  colData = coldata,
  design = ~condition
)
```

## --- Run DESeq2 ---
```{r}
# Run the DESeq pipeline
dds <- DESeq(dds)

# Save the DESeq2 object to an RDS file for future use
saveRDS(dds, file = "../data/DESeq2_object.rds")

# Get results
res <- results(dds)

# Optional: Order results by adjusted p-value
resOrdered <- res[order(res$padj), ]

# Summary of results
summary(res)

# Export results as a data frame
resDF <- as.data.frame(resOrdered)
resDF <- resDF %>% rownames_to_column("Ensembl_gene")
```

## --- Load Count matrix to extract gene names ---
```{r}
# Load file
cpm_matrix <- read.csv("../data/logCPM_for_plotting.csv", row.names = 1)

# Select gene names and fill with Ensembl in case of missing genes
gene_names <- cpm_matrix %>% select(gene) %>% 
  rownames_to_column("Ensembl_gene") %>% 
  mutate(gene = case_when(
  !str_detect(gene, "[A-Za-z0-9]") ~ Ensembl_gene,
  TRUE ~ gene
))
finaldf <- left_join(resDF, gene_names, by = "Ensembl_gene") %>% 
  mutate(gene = case_when(
  is.na(gene) ~ Ensembl_gene,
  TRUE ~ gene
))

# Save results to a CSV file
finaldf
write.csv(finaldf, file = "../data/DESeq2_results.csv")
```


#Plotting the HeatMap
## --- Regularized Log (rlog) Transformation:
```{r}
# Apply rlog transformation (ideal for visualization)
rlog_counts <- rlog(dds, blind = TRUE)

# Save the rlog-transformed counts to a file
saveRDS(rlog_counts, file = "rlog_transformed_counts.rds")
```


```{r}
# Load file
rlog_counts <- readRDS("../data/rlog_transformed_counts.rds")

# Extract matrix for heatmap
heatmap_matrix <- assay(rlog_counts)
```


#Select Specific Genes for Heatmap
```{r}
# Specify genes of interest
# Load resDF
resDF <- read.csv("../data/DESeq2_results.csv")

# Select top n genes to plot
n <- 50

# Pull up and down regulated genes
top_up <- resDF %>%
  arrange(padj) %>%
  filter(log2FoldChange > 0) %>%
  slice_head(n = n) %>%
  pull(Ensembl_gene)
top_down <- resDF %>%
  arrange(padj) %>%
  filter(log2FoldChange < 0) %>%
  slice_head(n = n) %>%
  pull(Ensembl_gene)
genes_of_interest <- c(top_up, top_down) 

# Ensure gene names match your rownames in heatmap_matrix
selected_genes <- heatmap_matrix[rownames(heatmap_matrix) %in% genes_of_interest, ]

idx <- match(rownames(selected_genes), resDF$Ensembl_gene)

all(resDF[idx, "Ensembl_gene"] == rownames(selected_genes))

#Rename matrix with matched genes
rownames(selected_genes) <- resDF[idx, "gene"]

# Transpose and scale for visualization
scaled_heatmap_matrix <- t(scale(t(selected_genes)))
```

```{r}
library(pheatmap)
# Save as a vector PDF
pdf(paste0("../output/Supp N - Figures and Analysis/figures/Supp_N_DeepVsSQ_top", n, "_DGE_heatmap.pdf"), width = 6, height = 8) # Adjust size in inches
pheatmap(
  scaled_heatmap_matrix,
  cluster_rows = TRUE, # Remove row clustering (no dendrogram for rows)
  cluster_cols = FALSE, # Remove column clustering (no dendrogram for columns)
  show_rownames = TRUE, # Hide row names
  fontsize_row = 6 / (n / 50), # Reduce row name font size (adjust value as needed)
  color = colorRampPalette(c("blue", "white", "red"))(50),
  main = paste("Top", n, "DGE - Deep vs Subcutaneous")
)
dev.off() # Close the PDF device


pheatmap(
  scaled_heatmap_matrix,
  cluster_rows = TRUE, # Remove row clustering (no dendrogram for rows)
  cluster_cols = FALSE, # Remove column clustering (no dendrogram for columns)
  show_rownames = TRUE, # Hide row names
  fontsize_row = 6 / (n / 50), # Reduce row name font size (adjust value as needed)
  color = colorRampPalette(c("blue", "white", "red"))(50),
  main = paste("Top", n, "DGE - Deep vs Subcutaneous")
)
```

# Plotting the Genes Differences between Deep and SC
## --- Select the genes ---
```{r}
# Specify genes of interest
genes_of_interest <- c(
  "TSHR", # TSH receptor
  "THRA", # Thyroid hormone receptor alpha
  "THRB", # Thyroid hormone receptor beta
  "SLC16A10",
  "SLC16A2",
  "UCP1", # Uncoupling protein 1
  "ADRB3", # Beta-3 adrenergic receptor
  "PRDM16",
  "EBF2",
  "PPARGC1A",
  "DIO1", # Deiodinase 1
  "DIO2", # Deiodinase 2
  "DIO3", # Deiodinase 3
  "CKMT1A",
  "CKMT2",
  "CKB",
  "SLC6A8",
  "ALPL"
)

genes_of_interest <- genes_of_interest[genes_of_interest %in% resDF$gene]
```

## --- Normalize the counts and log transform
```{r}
# Load the DESEQ2 object
dds <- readRDS("../data/DESeq2_object.rds")

# Extract normalized counts from DESeq2 object
norm_counts <- counts(dds, normalized = TRUE)
log10_norm_counts <- log10(norm_counts + 1)
norm_counts <- log10_norm_counts
```

## --- Filter the matrix ---
```{r}
# Restart resDF
resDF <- read.csv("../data/DESeq2_results.csv")

ens_genes_of_interest <- resDF %>%
  dplyr::filter(gene %in% genes_of_interest) %>%
  pull(Ensembl_gene)

# Ensure gene names match your rownames in norm_counts
selected_norm_counts <- norm_counts[rownames(norm_counts) %in% ens_genes_of_interest, ]
idx <- match(rownames(selected_norm_counts), resDF$Ensembl_gene)
all(resDF[idx, "Ensembl_gene"] == rownames(selected_norm_counts))

rownames(selected_norm_counts) <- resDF[idx, "gene"]

# Melt data for plotting (reshape to long format)
norm_counts_long <- melt(as.data.frame(selected_norm_counts), variable.name = "Sample", value.name = "Expression")

# Add metadata
norm_counts_long <- merge(norm_counts_long, coldata, by.x = "Sample", by.y = "row.names") %>% mutate(
  gene = rep(rownames(selected_norm_counts), times = nrow(norm_counts_long) / length(genes_of_interest)),
  subject_id = str_extract(Sample, "\\d+")
)
```

## --- Plot and save genes individually ---
```{r}
#Initialize plot list
plots <- list()

#Start plotting loop
for (plot.gene in genes_of_interest) {
  # Get pvalues
  # Subset and reorder the dataframe
  p_values <- resDF %>%
    filter(gene == plot.gene) %>%
    select(gene, padj)

  # Merge p-values with the original data to get y-position for annotations
  annotated_data <- norm_counts_long %>%
    filter(gene == plot.gene) %>%
    group_by(gene, condition) %>%
    summarise(max_expression = max(Expression), .groups = "drop") %>%
    left_join(p_values, by = "gene") %>%
    group_by(gene) %>%
    summarise(
      p_value = unique(padj), # select DESeq2 adjusted p-value
      y_position = max(max_expression) + 0.5, # Adjust placement
      fdr_label = ifelse(p_value < 0.01, "FDR < 0.01", paste0("FDR = ", signif(p_value, 1)))
    )
  if (annotated_data$p_value < 0.05) {
    print(plot.gene)
  }
  plot.data <- norm_counts_long %>%
    filter(gene == plot.gene)

  # Add p-value annotations to the plot
  p <- ggplot(plot.data, aes(x = condition, y = Expression)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0) +
    geom_line(aes(group = subject_id), alpha = 1, linewidth = 0.3) +

    # Add p-value annotations
    geom_text(
      data = annotated_data,
      aes(x = 1.5, y = y_position, label = fdr_label),
      inherit.aes = FALSE, # Prevent ggplot from inheriting x/y aesthetics
      size = 3 # Adjusted font size for annotation
    ) +

    # Labels and theme
    labs(
      title = plot.gene,
      x = NULL,
      y = "Expression (log10 Normalized Counts)"
    ) +
    theme_classic() +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 8, hjust = 0.5, face = "bold"), # Adjusted font size for title
      axis.title.y = element_text(size = 7), # Adjusted font size for axis label
      axis.text = element_text(size = 6), # Adjusted font size for axis text
      strip.text = element_text(size = 7) # Adjusted font size for facet labels
    )

  print(p)

  plots[[plot.gene]] <- p

  # Save the plot
  ggsave(
    filename = paste0("./figures/Paper_figures/Figure_3/", plot.gene, ".png"),
    plot = p,
    device = "png",
    units = "in",
    width = 1.5,
    height = 2.25,
    dpi = 300
  )
}
```

# Plotting local T3 and local T4 in deep and SC
```{r}
# meta is loaded on setup
meta

# format meta
colnames(meta)[3:6]
if (any(colnames(meta) == "Tissue")) {
  meta <- meta %>% dplyr::rename("tissue" = "Tissue", "subject" = "Donor_ID")
  meta <- meta %>% dplyr::rename("subject" = "Donor_ID")
}

# Step 1: Collect raw p-values
pvals <- c()
variables <- colnames(meta)[3:6]

for (variable in variables) {
  test_result <- compare_means(as.formula(paste0(variable, "~ tissue")), meta, paired = TRUE, p.adjust.method = "none")
  pvals <- c(pvals, test_result$p)
}

# Step 2: Apply FDR correction
fdr_vals <- p.adjust(pvals, method = "fdr")

# Step 3: Generate plots with FDR values
for (i in seq_along(variables)) {
  variable <- variables[i]
  fdr <- fdr_vals[i]

  # Get y-position for annotation
  annotated_data <- meta %>%
    select(!!sym(variable), tissue) %>%
    filter(!is.na(!!sym(variable))) %>%
    group_by(tissue) %>%
    summarise(max_expression = max(!!sym(variable)), .groups = "drop") %>%
    summarise(
      y_position = max(max_expression) + 0.5
    ) %>%
    mutate(FDR = signif(fdr, digits = 1))

  # Define annotation
  hormone <- str_extract(variable, "T\\d")
  plot_title <- ifelse(hormone == "T3", "Tissue T3", "Tissue T4") # Detects if T3 or T4
  y_axis_title <- ifelse(str_detect(variable, "tissue"), paste0(hormone, "nM/mg tissue"), paste0(hormone, " nM/µg protein in tissue"))



  # Plot
  p <- ggplot(meta, aes(x = tissue, y = !!sym(variable))) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0) +
    geom_line(aes(group = subject), alpha = 1, size = 0.3) +
    geom_text(
      data = annotated_data,
      aes(x = 1.5, y = y_position, label = paste0("FDR = ", signif(FDR, 3))),
      inherit.aes = FALSE,
      size = 3
    ) +
    theme_classic() +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 8, hjust = 0.5, face = "bold"),
      axis.title.y = element_text(size = 7),
      axis.text = element_text(size = 6),
      strip.text = element_text(size = 7),
      axis.title.x = element_blank()
    ) +
    labs(
      title = plot_title,
      y = y_axis_title
    )

  print(p)

  plots[[variable]] <- p
  ggsave(
    filename = paste0("./figures/Paper_figures/Figure_3/", variable, ".png"),
    plot = p,
    device = "png",
    units = "in",
    width = 1.5,
    height = 2.25,
    dpi = 300
  )
}


pvals
fdr_vals
```
#Saving the panel


```{r}
variables

names(plots) <- gsub("\\.", "_", names(plots))
ordered_plot_names <- list(
  c("T3_μg_protein", "T4_μg_protein", "T3_mg_tissue", "T4_mg_tissue"),
  c("TSHR", "THRA", "THRB"),
  c("SLC16A2", "SLC16A10", "SLC6A8"),
  c("DIO1", "DIO2", "DIO3", "ADRB3"),
  c("UCP1", "PRDM16", "EBF2", "PPARGC1A"),
  c("CKB", "CKMT1A", "CKMT2", "ALPL")
)

max_cols <- max(sapply(ordered_plot_names, length))

# Flatten into a grid row-wise with NULL padding
flat_layout <- unlist(lapply(ordered_plot_names, function(row) {
  c(row, rep(NA, max_cols - length(row))) # pad to match row length
}), use.names = FALSE)

tagged_ordered_plots <- vector("list", length(flat_layout))
tag_counter <- 1

for (i in seq_along(flat_layout)) {
  gene <- flat_layout[i]

  if (!is.na(gene)) {
    # Add tag to the plot
    tagged_ordered_plots[[i]] <- plots[[gene]] + labs(tag = LETTERS[tag_counter])
    tag_counter <- tag_counter + 1
  } else {
    # Insert blank
    tagged_ordered_plots[[i]] <- patchwork::plot_spacer()
  }
}

library(patchwork)

final_plot <- wrap_plots(tagged_ordered_plots, ncol = max_cols)

ggsave("./figures/Paper_figures/Figure_3/DESeq_Panel.png", plot = final_plot, width = 4, height = 8, scale = 1.7, units = "in", dpi = 600, bg = "whitesmoke")

ggsave("../output/2 - Figures and Analysis/figures/Figure_2_deseq_genes.png", plot = final_plot, width = 4, height = 8, scale = 1.7, units = "in", dpi = 600, bg = "white")

ggsave("../output/2 - Figures and Analysis/figures/Figure_2_deseq_genes.tiff", plot = final_plot, width = 4, height = 8, scale = 1.7, units = "in", dpi = 600, bg = "white")

pdf("../output/2 - Figures and Analysis/figures/Figure_2_deseq_genes.pdf", width = 4*1.7, height = 8*1.7)
final_plot
dev.off()

ggsave("../output/2 - Figures and Analysis/figures/Figure_2_deseq_genes.pdf",
       plot = final_plot,
       width = 4, height = 8, scale = 1.7, units = "in", bg = "white")

```


