---
title: "Gavrila_reactome_summary"
author: "Henrique Camara"
date: '2024-06-13'
output: html_document
---

```{r setup, include=FALSE}
# Set global chunk options for knitr
knitr::opts_chunk$set(echo = TRUE)

# Load necessary libraries
library(ggplot2)
library(tidyverse)
library(stringr)
library(openxlsx)

# Define a custom theme
custom_theme <- theme_bw() + theme(
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  plot.title = element_text(size = 16, face = "bold"),
  legend.title = element_text(size = 14),
  legend.text = element_text(size = 12),
  strip.text = element_text(size = 14)
)

# Apply the theme globally
theme_set(custom_theme)


# Alternative text wrapping function
wrap_text <- function(x, width) {
  str_wrap(x, width = width)
}

# Function to format text: uppercase and replace special characters
format_text <- function(text) {
  # Convert text to uppercase
  text_upper <- str_to_upper(text)

  # Replace special characters with underscores
  text_clean <- str_replace_all(text_upper, "[^A-Z0-9]", "_")

  # Remove potential double underscores
  text_clean <- str_replace_all(text_clean, "__+", "_")

  return(text_clean)
}

# Function to convert to sentence case
sentence_case <- function(text) {
  # Convert the first letter to uppercase and the rest to lowercase
  paste0(toupper(substr(text, 1, 1)), tolower(substr(text, 2, nchar(text))))
}
```

# Paper figures
The analysis above was useful to summarize data.
After discussion we decided to use the adjusted analysis as our main reference point. Thus, plots will be centered towards that.
We can check consistency in the other adjustments.

To create the paper figures, I plan to get the top 10 Positive and negative associated pathways and plot them as a DotPlot, with NES on x-axis, Cluster in the Y-axis, color as p.value

# --- Reading BMI adjusted

```{r}
path_list <- list()
# Define a function to format text (ensure this is defined elsewhere in your code)
format_text <- function(text) {
  # Convert text to uppercase
  text_upper <- str_to_upper(text)
  
  # Replace special characters with underscores
  text_clean <- str_replace_all(text_upper, "[^A-Z0-9]", "_")
  text_clean <- gsub("__", "_", text_clean)  # Replace potential double underscores created
  
  return(text_clean)
}

# Path to the Excel file
path <- "../data/neck_fat_rnaseq_adj_covars/reactome_gsea.xlsx"

sheet_names <- getSheetNames(path)

sheets <- lapply(sheet_names, function(sheet) {
  df <- read.xlsx(path, sheet = sheet)
  df$Cluster <- sheet
  df$Description <- format_text(df$Description)
  return(df)
})

read.xlsx(path)
names(sheets) <- sheet_names
BMI_temp <- bind_rows(sheets)
path_list[["BMI_temp"]] <- BMI_temp
rm(BMI_temp)

```

```{r}
# Load only the data form the covariate corrected files
covar_data <- path_list[["BMI_temp"]]

# Replace the pahtway name that is too big
covar_data <-
  covar_data %>% mutate(Description = case_when(
    Description == "RESPIRATORY_ELECTRON_TRANSPORT_ATP_SYNTHESIS_BY_CHEMIOSMOTIC_COUPLING_AND_HEAT_PRODUCTION_BY_UNCOUPLING_PROTEINS_" ~ "CHEMIOSMOTIC_COUPLING_AND_HEAT_PRODUCTION_BY_UNCOUPLING_PROTEINS",
    TRUE ~ Description
  ))
```

```{r}
grep("creatine", covar_data$Description, ignore.case = T, value = T)
```

Plot the data - Top 20 genes
```{r}
library(ggplot2)
library(dplyr)
library(stringr)

# Define date prefix
date_prefix <- format(Sys.Date(), "%Y-%m-%d")

# Define number of top genes
n_top <- 20

# Get unique covariates and exclude specific ones
covars <- unique(covar_data$Cluster)
covars <- covars[!grepl("Deep_and_SC", covars)] # Remove the Deep and SC analysis

# Create output directory if it does not exist
output_dir <- paste0("./figures/Paper_figures/DotPlot/top_", n_top)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Create PDF to save plots with date prefix
pdf(file.path("./figures/", paste0(date_prefix, "_GSEA_summary_top_padj.pdf")), width = 12, height = 6)

# Loop through each covariate and generate plots
for (covar in covars) {
  plot.data <- covar_data %>% filter(Cluster == covar)

  # Order with positives first
  pos_nes <- plot.data %>%
    filter(NES > 0) %>%
    arrange(p.adjust) %>%
    slice_head(n = n_top)
  neg_nes <- plot.data %>%
    filter(NES < 0) %>%
    arrange(desc(p.adjust)) %>%
    slice_tail(n = n_top)
  plot.data <- bind_rows(pos_nes, neg_nes)

  # Calculate min and max for -log10(p.adjust)
  min_neg_log_padj <- min(-log10(plot.data$p.adjust))
  max_neg_log_padj <- max(-log10(plot.data$p.adjust))

  # Create plot
  g <- ggplot(plot.data, aes(NES, Description)) +
    geom_count(aes(size = -log10(p.adjust), fill = NES), shape = 21, color = "black", stroke = 0.5) +
    scale_y_discrete(labels = function(x) wrap_text(x, width = 10)) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    scale_y_discrete(limits = rev(unique(plot.data$Description))) +
    theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
    labs(title = covar)

  # Print plot to PDF
  print(g)

  # Save plot as PNG with date prefix
  # ggsave(
  #   filename = file.path(output_dir, paste0(date_prefix, "_", covar, "_GSEA_summary_top_padj.png")),
  #   plot = g, device = "png", width = 12, height = 6, units = "in", dpi = 300
  # )
}

# Close the PDF device
dev.off()
```
### DotPlots - SC and Deep side by side

```{r}
# Get the pathways from the file to look more closely, potentially using LLM to filter the thermogenesis related paths
reactome_descriptions <- unique(covar_data$Description)
write.csv(reactome_descriptions, file = "./figures/Paper_figures/Reactome_description_list.csv", row.names = F)

# Assign thermogenic pathways to a vector
thermo_paths <- c(
  # These are bioenergetics/thermogenesis-related pathways
  "CHEMIOSMOTIC_COUPLING_AND_HEAT_PRODUCTION_BY_UNCOUPLING_PROTEINS", "THE_CITRIC_ACID_TCA_CYCLE_AND_RESPIRATORY_ELECTRON_TRANSPORT", "RESPIRATORY_ELECTRON_TRANSPORT_ATP_SYNTHESIS_BY_CHEMIOSMOTIC_COUPLING_AND_HEAT_PRODUCTION_BY_UNCOUPLING_PROTEINS", "RESPIRATORY_ELECTRON_TRANSPORT", "FATTY_ACID_METABOLISM", "PYRUVATE_METABOLISM_AND_CITRIC_ACID_TCA_CYCLE", "INTEGRATION_OF_ENERGY_METABOLISM", "REGULATION_OF_LIPID_METABOLISM_BY_PPARALPHA", "KETONE_BODY_METABOLISM", "REGULATION_OF_GLYCOLYSIS_BY_FRUCTOSE_2_6_BISPHOSPHATE_METABOLISM", "INTRACELLULAR_METABOLISM_OF_FATTY_ACIDS_REGULATES_INSULIN_SECRETION",

  # These are mitochondrial related pathways
  "MITOCHONDRIAL_TRANSLATION", "MITOCHONDRIAL_TRANSLATION_ELONGATION", "MITOCHONDRIAL_TRANSLATION_INITIATION", "MITOCHONDRIAL_TRANSLATION_TERMINATION", "MITOCHONDRIAL_PROTEIN_IMPORT", "MITOCHONDRIAL_BIOGENESIS", "MITOCHONDRIAL_FATTY_ACID_BETA_OXIDATION", "BIOLOGICAL_OXIDATIONS", "MITOCHONDRIAL_FATTY_ACID_BETA_OXIDATION_OF_UNSATURATED_FATTY_ACIDS", "MITOCHONDRIAL_UNCOUPLING",

  # This is creatine metabolism related
  "CREATINE_METABOLISM"
)
```

```{r}
# Load only the data form the covariate corrected files
covar_data <- path_list[["BMI_temp"]]

# Replace the pahtway name that is too big
covar_data <-
  covar_data %>% mutate(Description = case_when(
    Description == "RESPIRATORY_ELECTRON_TRANSPORT_ATP_SYNTHESIS_BY_CHEMIOSMOTIC_COUPLING_AND_HEAT_PRODUCTION_BY_UNCOUPLING_PROTEINS_" ~ "CHEMIOSMOTIC_COUPLING_AND_HEAT_PRODUCTION_BY_UNCOUPLING_PROTEINS",
    TRUE ~ Description
  ))

names(covar_data)
```


Plot the data
```{r}
# Define date prefix
date_prefix <- format(Sys.Date(), "%Y-%m-%d")

# Format the data to have the covariate independent of tissue location
covar_data <- covar_data %>%
  mutate( # Extract text before the dot
    "Tissue" = str_extract(Cluster, "^[^.]+"),
    # ^[^.]+ means match the beginning of the string (^) and then match one or more characters that are not a dot ([^.]+)
    # Extract text after the dot
    "Covariate" = str_extract(Cluster, "(?<=\\.)[^.]+")
    # (?<=\\.) means match only the text that follows a dot (lookbehind assertion), and [^.]+ matches one or more characters that are not a dot
  ) %>%
  mutate(
    Tissue = case_when(
      Tissue == "SC" ~ "SubQ",
      TRUE ~ Tissue # Keep existing value if not matched
    )
  ) %>%
  filter(Tissue != "Deep_and_SC")


# Create a data frame with all possible combinations of Description, Tissue and Covariate
all_combinations <- expand.grid(
  Description = unique(covar_data$Description),
  Tissue = unique(covar_data$Tissue),
  Covariate = unique(covar_data$Covariate),
  stringsAsFactors = FALSE
)

# Merge with the original data, replacing missing values
f.covar_data <- all_combinations %>%
  left_join(covar_data, by = c("Description", "Tissue", "Covariate")) %>%
  mutate(
    NES = ifelse(is.na(NES), 0, NES),
    p.adjust = ifelse(is.na(p.adjust), 1, p.adjust)
  )

unique(f.covar_data$Covariate)


# Get unique covariates and exclude the ones combining DEEP and SC for the correlation
covars <- f.covar_data %>%
  filter(Tissue != "Deep_and_SC") %>%
  pull(Covariate) %>%
  unique()
```

#Rename the covariates data
```{r}
# f.covar_data = f.covar_data %>% mutate(Covariate = case_when(
#                                                     Covariate == "FreeT4" ~ "Free T4",
#                                                     Covariate == "T4_per_mcg_protein" ~ "T4/μg of protein",
#                                                     Covariate == "T3_per_mcg_protein" ~ "T3/μg of protein",
#                                                     Covariate == "T4_per_mg_tissue" ~ "T4/mg of tissue",
#                                                     Covariate == "T3_per_mg_tissue" ~ "T3/mg of tissue",
#                                                     Covariate == "FreeT3" ~ "Free T3",
#                                                     Covariate == "HYPERvsHYPO" ~ "Hyperthyroidism",
#                                                     Covariate == "TSH" ~ "TSH"))
```


#### Plotting thermogenic pathways==================================



```{r}
# Define widht and heigth of plots
w <- 12
h <- 5

# Create output directory if it does not exist
output_dir <- paste0("./figures/Paper_figures/DotPlot/DEEP_SC_Thermogenic_pathways")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

#### Plotting thermogenic pathways==================================
# Create PDF to save plots with date prefix
# pdf(file.path(output_dir, paste0(date_prefix, "_GSEA_summary_thermo.pdf")), width = 12, height = 6)


# Loop through each covariate and generate plots
for (covar in covars) {
  plot.data <- f.covar_data %>% filter(
    Covariate == covar,
    Description %in% thermo_paths
  ) # Filter for the thermogenic pathways

  # Order with positives first
  pos_nes <- plot.data %>%
    filter(NES >= 0) %>%
    arrange(p.adjust)
  neg_nes <- plot.data %>%
    filter(NES < 0) %>%
    arrange(desc(p.adjust))

  plot.data <- bind_rows(pos_nes, neg_nes)

  # Calculate min and max for -log10(p.adjust)
  min_neg_log_padj <- min(-log10(plot.data$p.adjust))
  max_neg_log_padj <- max(-log10(plot.data$p.adjust))

  # Create plot ordered by NES
  g <- ggplot(plot.data, aes(Tissue, Description)) +
    geom_count(aes(size = -log10(p.adjust), fill = NES), shape = 21, color = "black", stroke = 0.5) +
      # Add asterisks for p.adjust < 0.05
  geom_text(
    data = plot.data %>% filter(p.adjust < 0.05),
    aes(label = "*"),
    color = "black",
    size = 6,
    vjust = 0.5,
    hjust = 1.75
  ) + # Adjust the position of the asterisk
    scale_y_discrete(labels = function(x) wrap_text(x, width = 10)) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    scale_y_discrete(limits = rev(unique(plot.data$Description))) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14), # Increase x-axis text size
      axis.text.y = element_text(size = 14), # Increase y-axis text size
      axis.title.x = element_blank(), # Remove x-axis title
      axis.title.y = element_blank(), # Remove y-axis title
      plot.title = element_text(size = 16, face = "bold"), # Increase title size and make it bold
      legend.text = element_text(size = 12), # Increase legend text size
      legend.title = element_text(size = 14) # Increase legend title size
    ) +
    labs(title = paste0(covar, "\n", "Thermogenesis-related pathways"))

  # Print plot to PDF
  print(g)

  # Save plot as PNG with date prefix
  ggsave(
    filename = file.path(output_dir, "NES_ordered_Y", paste0(date_prefix, "_", covar, "_GSEA_thermogenic_paths.png")),
    plot = g, device = "png", width = w, height = h, units = "in", dpi = 300, create.dir = T
  )
  
  ggsave(
    filename = file.path("../output/5_6 - Figures and Analysis/figures",  paste0("Figure_N_", date_prefix, "_", covar, "_thermogenic_pathways.png")),
    plot = g, device = "png", width = w, height = h, units = "in", dpi = 600, create.dir = T
  )
  
  pdf(file.path("../output/5_6 - Figures and Analysis/figures",  paste0("Figure_N_", date_prefix, "_", covar, "_thermogenic_pathways.pdf")), width = w, height = h)


  # Create plot ordered alphabetically
  g <- ggplot(plot.data, aes(Tissue, Description)) +
    geom_count(aes(size = -log10(p.adjust), fill = NES), shape = 21, color = "black", stroke = 0.5) +
    scale_y_discrete(labels = function(x) wrap_text(x, width = 10)) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14), # Increase x-axis text size
      axis.text.y = element_text(size = 14), # Increase y-axis text size
      axis.title.x = element_blank(), # Remove x-axis title
      axis.title.y = element_blank(), # Remove y-axis title
      plot.title = element_text(size = 16, face = "bold"), # Increase title size and make it bold
      legend.text = element_text(size = 12), # Increase legend text size
      legend.title = element_text(size = 14) # Increase legend title size
    ) +
    labs(title = paste0(covar, "\n", "Thermogenesis-related pathways"))

  # Print plot to PDF
  print(g)
  # Save plot as PNG with date prefix
  # ggsave(
  #   filename = file.path(output_dir, "Alphabetical_Y", paste0(date_prefix, "_", covar, "_GSEA_thermogenic_paths.png")),
  #   plot = g, device = "png", width = w, height = h, units = "in", dpi = 300, create.dir = T
  # )
}

# Close the PDF device
# dev.off()
```
#### Plot all covariates side by side
```{r}
# Define widht and heigth of plots
w <- 12
h <- 5

# Create output directory if it does not exist
output_dir <- paste0("./figures/Paper_figures/DotPlot/DEEP_SC_Thermogenic_pathways/ALL")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

#### Plotting thermogenic pathways==================================
# # Create PDF to save plots with date prefix
# pdf(file.path(output_dir, paste0(date_prefix, "_GSEA_summary_thermo.pdf")), width = 12, height = 6)


# Loop through each covariate and generate plots
plot.data <- f.covar_data %>% filter(Description %in% c(thermo_paths)) # Filter for the thermogenic pathways
# head(plot.data)

plot.data <- plot.data %>%
  filter(!Covariate %in% c("T4_per_mcg_protein", "T3_per_mcg_protein")) %>%
  mutate(Covariate = case_when(
    Covariate == "FreeT4" ~ "Serum Free T4",
    Covariate == "T4_per_mcg_protein" ~ "T4/μg of protein",
    Covariate == "T3_per_mcg_protein" ~ "T3/μg of protein",
    Covariate == "T4_per_mg_tissue" ~ "T4/mg of tissue",
    Covariate == "T3_per_mg_tissue" ~ "T3/mg of tissue",
    Covariate == "FreeT3" ~ "Serum Free T3",
    Covariate == "HYPERvsHYPO" ~ "Hyperthyroidism vs Hypothyroidism",
    Covariate == "TSH" ~ "Serum TSH"
  ))

# Order with positives first
pos_nes <- plot.data %>%
  filter(NES >= 0) %>%
  arrange(p.adjust)
neg_nes <- plot.data %>%
  filter(NES < 0) %>%
  arrange(desc(p.adjust))

plot.data <- bind_rows(pos_nes, neg_nes)

# Calculate min and max for -log10(p.adjust)
min_neg_log_padj <- min(-log10(plot.data$p.adjust))
max_neg_log_padj <- max(-log10(plot.data$p.adjust))

# Create plot ordered by NES
g <- ggplot(plot.data, aes(Covariate, Description)) +
  geom_count(aes(size = -log10(p.adjust), fill = NES, color = Tissue), shape = 21, stroke = 0.5) +
  scale_y_discrete(labels = function(x) wrap_text(x, width = 10)) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_y_discrete(limits = rev(unique(plot.data$Description))) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14), # Increase x-axis text size
    axis.text.y = element_text(size = 14), # Increase y-axis text size
    axis.title.x = element_blank(), # Remove x-axis title
    axis.title.y = element_blank(), # Remove y-axis title
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5), # Increase title size and make it bold
    legend.text = element_text(size = 12), # Increase legend text size
    legend.title = element_text(size = 14) # Increase legend title size
  ) +
  labs(title = paste0("Thermogenesis-related pathways"))


# GPT code
# Create plot ordered by NES
g <- ggplot(plot.data, aes(Tissue, Description)) +
  geom_count(aes(size = -log10(p.adjust), fill = NES), shape = 21, stroke = 0.5) +
  facet_grid(. ~ Covariate, scales = "free_x", space = "free_x") + # Facet by Tissue with separate x-axis
  # Add asterisks for p.adjust < 0.05
  geom_text(
    data = plot.data %>% filter(p.adjust < 0.05),
    aes(label = "*"),
    color = "black",
    size = 6,
    vjust = 0
  ) + # Adjust the position of the asterisk
  scale_y_discrete(labels = function(x) wrap_text(x, width = 10)) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_y_discrete(limits = rev(unique(plot.data$Description))) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14), # Increase x-axis text size
    axis.text.y = element_text(size = 14), # Increase y-axis text size
    axis.title.x = element_blank(), # Remove x-axis title
    axis.title.y = element_blank(), # Remove y-axis title
    plot.title = element_text(size = 16, face = "bold"), # Increase title size and make it bold
    legend.text = element_text(size = 12), # Increase legend text size
    legend.title = element_text(size = 14), # Increase legend title size
    strip.text.x = element_text(size = 12, face = "bold") # Style for facet labels (Tissue)
  ) +
  labs(title = paste0("Thermogenesis-related pathways"))



############################
# Order Covariate
desired_order <- c("Serum TSH", "Serum Free T4", "Serum Free T3", "T4/mg of tissue", "T3/mg of tissue", "Hyperthyroidism vs Hypothyroidism") # Replace with your actual facet levels
plot.data$Covariate <- factor(plot.data$Covariate, levels = desired_order)

plot.data$Description <- gsub("_", " ", plot.data$Description)

plot.data <- plot_data %>% filter(Covariate %in% desired_order)

# Create plot ordered by NES
g <- ggplot(plot.data, aes(Tissue, Description)) +
  geom_count(aes(size = -log10(p.adjust), fill = NES), shape = 21, stroke = 0.5) +
  facet_grid(. ~ Covariate,
    scales = "free_x", space = "free_x",
    labeller = labeller(Covariate = function(x) str_wrap(x, width = 15))
  ) + # Wrap facet labels
  # Add asterisks for p.adjust < 0.05
  geom_text(
    data = plot.data %>% filter(p.adjust < 0.05),
    aes(label = "*"),
    color = "black",
    size = 6,
    vjust = 0
  ) + # Adjust the position of the asterisk
  scale_y_discrete(labels = function(x) str_wrap(x, width = 10)) + # Wrap y-axis labels
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_y_discrete(limits = rev(unique(plot.data$Description))) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14), # Increase x-axis text size
    axis.text.y = element_text(size = 14), # Increase y-axis text size
    axis.title.x = element_blank(), # Remove x-axis title
    axis.title.y = element_blank(), # Remove y-axis title
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5), # Center and increase title size
    legend.text = element_text(size = 12), # Increase legend text size
    legend.title = element_text(size = 14), # Increase legend title size
    strip.text.x = element_text(size = 12, face = "bold", margin = margin(b = 10, t = 10)) # Style for facet labels with spacing
  ) +
  labs(title = "Thermogenesis-related pathways")

# Print plot to PDF
print(g)

# Save plot as PNG with date prefix
ggsave(
  filename = file.path(output_dir, "NES_ordered_Y", paste0(date_prefix, "_", covar, "_GSEA_thermogenic_paths.png")),
  plot = g, device = "png", width = 20, height = 15, units = "in", dpi = 300, create.dir = T
)

# Save plot as PNG with date prefix
ggsave(
  filename = file.path("../output/4 - Figures and Analysis/figures/Figure_4_thermogenesis_all_covars_dotplot.png"),
  plot = g, device = "png", width = 20, height = 15, units = "in", dpi = 600, create.dir = T
)

pdf(file.path("../output/4 - Figures and Analysis/figures/Figure_4_thermogenesis_all_covars_dotplot.pdf"), width = 20, height = 15)
# Print plot to PDF
print(g)
# Close the PDF device
dev.off()
```


#### Plotting the top N genes pathways==================================
```{r}
# Define widht and heigth of plots
w <- 15
h <- 5

# Define top X
n_top <- 5

# Create output directory if it does not exist
output_dir <- paste0("./figures/Paper_figures/DotPlot/top_", n_top)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

#### Plotting thermogenic pathways==================================
# Create PDF to save plots with date prefix
pdf(file.path(output_dir, paste0(date_prefix, "_GSEA_summary_top", n_top, ".pdf")), width = 12, height = 6)


# Loop through each covariate and generate plots
for (covar in covars) {
  plot.data <- f.covar_data %>% filter(Covariate == covar)
  # plot.data$Description <- sentence_case(plot.data$Description)

  # Order with positives first
  pos_nes <- plot.data %>%
    filter(NES >= 0) %>%
    group_by(Tissue) %>%
    arrange(p.adjust) %>%
    slice_head(n = n_top)
  neg_nes <- plot.data %>%
    filter(NES < 0) %>%
    group_by(Tissue) %>%
    arrange(desc(p.adjust)) %>%
    slice_tail(n = n_top)

  plot.data <- bind_rows(pos_nes, neg_nes)

  # Calculate min and max for -log10(p.adjust)
  min_neg_log_padj <- min(-log10(plot.data$p.adjust))
  max_neg_log_padj <- max(-log10(plot.data$p.adjust))

  # Create plot ordered by NES
  g <- ggplot(plot.data, aes(Tissue, Description)) +
    geom_count(aes(size = -log10(p.adjust), fill = NES), shape = 21, color = "black", stroke = 0.5) +
      # Add asterisks for p.adjust < 0.05
  geom_text(
    data = plot.data %>% filter(p.adjust < 0.05),
    aes(label = "*"),
    color = "black",
    size = 6,
    vjust = 0.5,
    hjust = 1.75
  ) + # Adjust the position of the asterisk
    scale_y_discrete(labels = function(x) wrap_text(x, width = 10)) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    scale_y_discrete(limits = rev(unique(plot.data$Description))) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14), # Increase x-axis text size
      axis.text.y = element_text(size = 14), # Increase y-axis text size
      axis.title.x = element_blank(), # Remove x-axis title
      axis.title.y = element_blank(), # Remove y-axis title
      plot.title = element_text(size = 16, face = "bold"), # Increase title size and make it bold
      legend.text = element_text(size = 12), # Increase legend text size
      legend.title = element_text(size = 14) # Increase legend title size
    ) +
    labs(title = paste0(covar, "\n", "Top ", n_top, " correlated pathways"))

  # Print plot to PDF
  print(g)

  # Save plot as PNG with date prefix
  ggsave(
    filename = file.path(output_dir, "NES_ordered_Y", paste0(date_prefix, "_", covar, "_GSEA_thermogenic_paths.png")),
    plot = g, device = "png", width = w, height = h, units = "in", dpi = 300, create.dir = T
  )

  # Save plot as PNG with date prefix
  ggsave(
    filename = file.path("../output/5_6 - Figures and Analysis/figures",  paste0("Figure_N_", date_prefix, "_", covar, "_top_", n_top, "_pathways.png")),
    plot = g, device = "png", width = w, height = h, units = "in", dpi = 600, create.dir = T
  )
  
  pdf(file.path("../output/5_6 - Figures and Analysis/figures",  paste0("Figure_N_", date_prefix, "_", covar, "_top_", n_top, "_pathways.pdf")), width = w, height = h)
  print(g)
  dev.off()

  g_flipped <- g + coord_flip() + theme(axis.text.x = element_text(angle = 90))

  # Save plot as PNG with date prefix
  ggsave(
    filename = file.path(output_dir, "NES_ordered_Y", paste0(date_prefix, "_", covar, "_GSEA_thermogenic_paths_FLIPPED.png")),
    plot = g_flipped, device = "png", width = h, height = w, units = "in", dpi = 300, create.dir = T
  )
}

# Close the PDF device
dev.off()
```
```{r}
covar_data %>% filter(Description == "CREATINE_METABOLISM")
```

#### Plot all covariates side by side
```{r}
# Define widht and heigth of plots
w <- 12
h <- 5

# Create output directory if it does not exist
output_dir <- paste0("./figures/Paper_figures/DotPlot/DEEP_SC_Thermogenic_pathways/ALL")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

#### Plotting thermogenic pathways==================================
# # Create PDF to save plots with date prefix
# pdf(file.path(output_dir, paste0(date_prefix, "_GSEA_summary_thermo.pdf")), width = 12, height = 6)


# Loop through each covariate and generate plots
plot.data <- f.covar_data %>% filter(Description %in% thermo_paths) # Filter for the thermogenic pathways
# head(plot.data)

plot.data <- plot.data %>%
  filter(!Covariate %in% c("T4_per_mcg_protein", "T3_per_mcg_protein")) %>%
  mutate(Covariate = case_when(
    Covariate == "FreeT4" ~ "Serum Free T4",
    Covariate == "T4_per_mcg_protein" ~ "T4/μg of protein",
    Covariate == "T3_per_mcg_protein" ~ "T3/μg of protein",
    Covariate == "T4_per_mg_tissue" ~ "T4/mg of tissue",
    Covariate == "T3_per_mg_tissue" ~ "T3/mg of tissue",
    Covariate == "FreeT3" ~ "Serum Free T3",
    Covariate == "HYPERvsHYPO" ~ "Hyperthyroidism vs Hypothyroidism",
    Covariate == "TSH" ~ "Serum TSH"
  ))

# Order with positives first
pos_nes <- plot.data %>%
  filter(NES >= 0) %>%
  arrange(p.adjust)
neg_nes <- plot.data %>%
  filter(NES < 0) %>%
  arrange(desc(p.adjust))

plot.data <- bind_rows(pos_nes, neg_nes)

# Calculate min and max for -log10(p.adjust)
min_neg_log_padj <- min(-log10(plot.data$p.adjust))
max_neg_log_padj <- max(-log10(plot.data$p.adjust))

############################
# Order Covariate
desired_order <- c("Serum TSH", "Serum Free T4", "Serum Free T3", "T4/mg of tissue", "T3/mg of tissue", "Hyperthyroidism vs Hypothyroidism") # Replace with your actual facet levels
plot.data$Covariate <- factor(plot.data$Covariate, levels = desired_order)

plot.data$Description <- gsub("_", " ", plot.data$Description)
```

# --- Format plotting information
```{r}
threshold <- 0.05
# --- Filter for genes of interest
plot_data <- plot.data %>%
  mutate(
    significance = p.adjust < threshold,
    stroke_width = ifelse(significance, 0.5, 0.1),
    significance_label = ifelse(significance, paste0("FDR < ", threshold), "Not Significant"),
    significance_color = ifelse(significance_label == "Not Significant", "grey75", "black"),
    rounded_NES = round(NES, 2) # round to 2 decimal places

    ) 

# --- Determine nrows for size adjustment when saving
plot_nrow <- plot_data %>%
  distinct(Description) %>%
  nrow()

# --- Adjust stroke color
stroke_color <- plot_data %>%
  distinct(significance_label, significance_color) %>%
  deframe() # turns into a named vector (names are ensembl_ids)
```

```{r}
# Create plot ordered by NES
g <- ggplot(plot_data, aes(Tissue, Description)) +
  geom_point(
    aes(fill = NES, stroke = stroke_width, colour = significance_label),
    shape = 21, size = 3.5
  ) +
  geom_text(
    aes(label = rounded_NES),
    size = 3 / 2.81,
    color = "black",
    fontface = "bold"
  ) +
  scale_fill_gradient2(low = "royalblue", mid = "white", high = "firebrick2", midpoint = 0) +
  scale_color_manual(
    values = stroke_color, # Only include significant one
    name = "Significance"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    strip.background = element_blank()
  ) +
  theme(
    legend.position = "top",
    legend.key.size = unit(2, "mm"),
    legend.text = element_text(size = 4),
    legend.title = element_text(size = 5),
    axis.line = element_line(size = 0.3) # smaller = thinner lines
  ) 
g
g <- g + facet_wrap(vars(Covariate), ncol = length(unique(plot_data$Description)))
g +
  geom_count(aes(size = -log10(p.adjust), fill = NES), shape = 21, stroke = 0.5) +
  facet_grid(. ~ Covariate,
    scales = "free_x", space = "free_x",
    labeller = labeller(Covariate = function(x) str_wrap(x, width = 15))
  ) + # Wrap facet labels
  # Add asterisks for p.adjust < 0.05
  geom_text(
    data = plot.data %>% filter(p.adjust < 0.05),
    aes(label = "*"),
    color = "black",
    size = 6,
    vjust = 0
  ) + # Adjust the position of the asterisk
  scale_y_discrete(labels = function(x) str_wrap(x, width = 10)) + # Wrap y-axis labels
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_y_discrete(limits = rev(unique(plot.data$Description))) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14), # Increase x-axis text size
    axis.text.y = element_text(size = 14), # Increase y-axis text size
    axis.title.x = element_blank(), # Remove x-axis title
    axis.title.y = element_blank(), # Remove y-axis title
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5), # Center and increase title size
    legend.text = element_text(size = 12), # Increase legend text size
    legend.title = element_text(size = 14), # Increase legend title size
    strip.text.x = element_text(size = 12, face = "bold", margin = margin(b = 10, t = 10))   ) +# Style for facet labels with spacing

  labs(title = "Thermogenesis-related pathways")

# Print plot to PDF
print(g)

# Save plot as PNG with date prefix
ggsave(
  filename = file.path(output_dir, "NES_ordered_Y", paste0(date_prefix, "_", covar, "_GSEA_thermogenic_paths.png")),
  plot = g, device = "png", width = 20, height = 15, units = "in", dpi = 300, create.dir = T
)

# Save plot as PNG with date prefix
ggsave(
  filename = file.path("../output/temp/chunk16_Pathway_corrs.png"),
  plot = g, device = "png", width = 20, height = 15, units = "in", dpi = 300, create.dir = T
)

# # Close the PDF device
# dev.off()
```

# --- Plot
```{r}
p <- ggplot(plot_data, aes(x = var_labels, y = ensembl_id)) +
  geom_point(
    aes(fill = slope, stroke = stroke_width, colour = significance_label),
    shape = 21, size = 3.5
  ) +
  geom_text(
    aes(label = rounded_slope),
    size = 3 / 2.81,
    color = "black",
    fontface = "bold"
  ) +
  scale_fill_gradient2(low = "royalblue", mid = "white", high = "firebrick2", midpoint = 0) +
  scale_color_manual(
    values = stroke_color, # Only include significant one
    name = "Significance"
  ) +
  scale_y_discrete(labels = y_axis_text) +
  theme_pubr(
    base_size = 6,
    base_family = "Arial"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    strip.background = element_rect(fill = "white", color = "white")
  ) +
  theme(
    legend.position = "top",
    legend.key.size = unit(2, "mm"),
    legend.text = element_text(size = 4),
    legend.title = element_text(size = 5),
    axis.line = element_line(size = 0.3) # smaller = thinner lines
  ) +
  labs(
    y = NULL,
    x = NULL
  )

p <- p + facet_wrap(vars(tissue))
p
# Save plot
scale <- 95
ggsave(paste0("../output/4 - Figures and Analysis/figures/Figure_4.png"), create.dir = T, plot = p, device = "png", width = 1 * scale, height = (1 + 0.01 * plot_nrow) * scale, units = "mm", dpi = 600, bg = "white")
```


