---
title: "Gavrila_TSH_analysis"
author: "Henrique"
date: "2024-02-06"
output: html_document
---

```{r setup, include=FALSE}
library(openxlsx)
library(tidyverse)
library(ggpubr)

#Formatting the metadata
meta = read.xlsx("../data/Metadata_for_correlation.xlsx")

meta = meta %>% separate_wider_delim(cols = Sample, delim = " ", names = c("Tissue", "Donor_ID"))
meta = meta %>% rename("Serum_Free_T3" = "Free.T3",
                       "Serum_Free_T4" = "Free.T4",
                       "Serum_TSH" = "TSH",
                       "log10_Serum_TSH" = "log[TSH]"
                       ) %>%
  mutate("log10_Serum_TSH" = log10(Serum_TSH))
  


names(meta) = gsub("[ /]", "_", names(meta))


head(meta)
names(meta)

correct_mg_string <- function(x) {
  # First, replace '_mg' with '/mg'
  x <- gsub("_mg", "/mg", x)
  
  # Next, replace '_μg' with '/μg'
  x <- gsub("_μg", "/μg", x)
  
  # Next, replace 'T3_' with 'T3/'
  x <- gsub("T3_", "T3/", x)
  
    # Next, replace 'T4_' with 'T4/'
  x <- gsub("T4_", "T4/", x)
  
  # Finally, replace any remaining '_' with a space
  x <- gsub("_", " ", x)
  
  return(x)
}
```

# ---Plotting serum to serum variables
```{r}
#Extract column with serum measurements
serum_cols = grep("Serum", names(meta), value = T)
plot.meta = meta %>% select(serum_cols, Tissue)
plot.meta$Serum_Free_T3 = as.numeric(plot.meta$Serum_Free_T3)
```


```{r}
combinations <- data.frame(x = c("Serum_TSH", "Serum_TSH", "Serum_Free_T4"),
                           y = c("Serum_Free_T4", "Serum_Free_T3", "Serum_Free_T3"))

combinations <- data.frame(x = c("log10_Serum_TSH", "log10_Serum_TSH", "Serum_Free_T4"),
                           y = c("Serum_Free_T4", "Serum_Free_T3", "Serum_Free_T3"))

plot_list <- list()
# Plot with correlation labels in the top-left corner
for(n in 1:nrow(combinations)){
    var1 <- combinations$x[n]
    var2 <- combinations$y[n]
    sp <- ggscatter(plot.meta %>% filter(Tissue == "Deep"), 
                    x = var1, y = var2,
      color = "black",
      add = "reg.line", 
      conf.int = T) #Change this to TRUE if you want to add the Coef interval
    
    # max.data = plot.meta %>% select(all_of(c(var1, var2))) %>% filter(!is.na(!!sym(var1)), !is.na(!!sym(var2)))
    # max.y = max(max.data[[var2]], na.rm = T)
    # max.x = max(max.data[[var1]], na.rm = T)
    
    sp = sp + stat_cor(
      method = "spearman", 
      cor.coef.name = "rho", 
      hjust = -0.1, 
      vjust = 0, 
      size = 6/2.81,
      show.legend = F) + 
      
      ##The code below should be useless based on the theme_set above
      theme_pubr(base_size = 6) +
      # scale_y_continuous(limits = c(0, max.y*1.2)) +
      # scale_x_continuous(limits = c(0, max.x*1.05)) +
      theme(
        plot.title = element_text(size = 8, face = "bold", hjust = 0.5),
      ) +
      labs(title = paste(correct_mg_string(var1), correct_mg_string(var2), sep = " x "),
           x = correct_mg_string(var1),
           y = correct_mg_string(var2)
           ) 
    plot_list[[paste(var1, var2, sep = "x")]] <- sp

}

plot_list
```

Subjects 1, 3, 5, 9, 10 and 12 had their blood collected for measurement of thyroid hormones before surgery.
Therefore, correlations of circulating and local adipose tissue thyroid hormones is not paired.
Those subjects will be excluded from correlations with local thyroid hormones.


#PLOTTING THE DATA 
```{r}

#--------------
# FILTER FOR SELECTED PATIENTS WITH THYROID TEST AT TIME OF SURGERY
#--------------

#Filter for subjects with paired blood drawn and surgery
subjects_to_keep <- c(2,4,6,7,8,11,13,14,15,16)

names(meta) = gsub("[ /]", "_", names(meta))
plot.meta = meta %>% select(-`log[TSH]`)
plot.meta$Serum_Free_T3 = as.numeric(plot.meta$Serum_Free_T3)

plot.meta <- plot.meta %>% 
  filter(Donor_ID %in% as.character(tyroid_at_surgery_patients)) %>% #Filtering for selected patients
  select(1,2,Serum_TSH, T4_mg.tissue, T3_mg.tissue, Serum_Free_T4, Serum_Free_T3) #At this point I can also filter for the graphs I am plotting only

plot.meta.1 <- plot.meta
# Plot with correlation labels in the top-left corner
for(col in (3:ncol(plot.meta))){
  col2 = 3
  while(col2 <= ncol(plot.meta)){
    var1 = names(plot.meta)[col]
    var2 = names(plot.meta)[col2]
    
    sp <- ggscatter(plot.meta, x = var1, y = var2,
      color = "Tissue", palette = c("#E66100", "#5D3A9B"),
      add = "reg.line", 
      conf.int = T) #Change this to TRUE if you want to add the Coef interval
    
    max.data = plot.meta %>% select(all_of(c(var1, var2))) %>% filter(!is.na(!!sym(var1)), !is.na(!!sym(var2)))
    max.y = max(max.data[[var2]], na.rm = T)
    max.x = max(max.data[[var1]], na.rm = T)
    sp = sp + stat_cor(
      aes(color = Tissue),
      method = "spearman", cor.coef.name = "rho", 
      label.x = -Inf, label.y = c(max.y*1.1*1.05, max.y*1.1), hjust = -0.1, vjust = 0, size = 6,
      show.legend = F) + 
      
      ##The code below should be useless based on the theme_set above
      theme_classic() +
      theme(
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 18),
        plot.title = element_text(size = 24, face = "bold"),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 16)

      ) +
      labs(title = paste(correct_mg_string(var1), correct_mg_string(var2), sep = " x "),
           x = correct_mg_string(var1),
           y = correct_mg_string(var2)
           ) +
      scale_y_continuous(limits = c(0, max.y*1.2)) +
      scale_x_continuous(limits = c(0, max.x*1.05))
    


    print(sp)
    # ggsave(create.dir = T, 
    #        filename = file.path("./figures/Paper_figures/Figure_4/Tyroid_at_surgery_filter/", var1, paste0(var1, "_x_", var2, ".png")),
    #       device = "png", 
    #       plot = sp,
    #       width = 6, height = 6,
    #       dpi = 600)
    col2 = col2 + 1
  }
}

names(plot.meta)
grep("u", names(plot.meta))
```



#Plotting serum to serum variables with Alina`s raw data

```{r}
meta <- read.xlsx("../data/BAT Collection Study_Gavrila_11_2022_Joslin.xlsx")
head(meta)

meta <- meta %>% filter(!is.na(`Study.ID#`)) %>% 
  select(`Study.ID#`,`Free.T4`,  `Free.T3`, `TSH`) %>% 
  rename("Serum_Free_T3" = "Free.T3",
                       "Serum_Free_T4" = "Free.T4",
                       "Serum_TSH" = "TSH")

meta
```

```{r}
names(meta) = gsub("[ /]", "_", names(meta))
serum_cols = grep("Serum", names(meta), value = T)
plot.meta = meta %>% select("Study.ID#", serum_cols)
plot.meta$Serum_Free_T3 = as.numeric(plot.meta$Serum_Free_T3)

plot.meta <- plot.meta %>% filter(!is.na(Serum_Free_T4), !is.na(Serum_TSH), !is.na(Serum_Free_T3))

# Plot with correlation labels in the top-left corner
for(col in (2:4)){
  col2 = 2
  while(col2 <= 4){
    var1 = names(plot.meta)[col]
    var2 = names(plot.meta)[col2]
    
    
    
    sp <- ggscatter(plot.meta, x = var1, y = var2,
      color = "black",
      add = "reg.line", 
      conf.int = T) #Change this to TRUE if you want to add the Coef interval
    
    max.data = plot.meta %>% select(all_of(c(var1, var2))) %>% filter(!is.na(!!sym(var1)), !is.na(!!sym(var2)))
    max.y = max(max.data[[var2]], na.rm = T)
    max.x = max(max.data[[var1]], na.rm = T)
    sp = sp + stat_cor(
      method = "spearman", cor.coef.name = "rho", 
      label.x = -Inf, label.y = c(max.y*1.1*1.05, max.y*1.1), hjust = -0.1, vjust = 0, size = 6,
      show.legend = F) + 
      
      ##The code below should be useless based on the theme_set above
      theme_classic() +
      theme(
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 18),
        plot.title = element_text(size = 24, face = "bold"),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 16)

      ) +
      labs(title = paste(correct_mg_string(var1), correct_mg_string(var2), sep = " x "),
           x = correct_mg_string(var1),
           y = correct_mg_string(var2)
           ) +
      scale_y_continuous(limits = c(0, max.y*1.2)) +
      scale_x_continuous(limits = c(0, max.x*1.05))
    


    print(sp)
    # ggsave(create.dir = T, 
    #        filename = file.path("./figures/Paper_figures/Figure_4/correlation_plots", var1, paste0(var1, "_x_", var2, ".png")),
    #       device = "png", 
    #       plot = sp,
    #       width = 6, height = 6,
    #       dpi = 600)
    col2 = col2 + 1
  }
}

names(plot.meta)
grep("u", names(plot.meta))
```

```{r}
plot.meta
plot.meta.1

all(plot.meta$Serum_TSH == plot.meta.1$Serum_TSH)
```

```{r}
#--------------
# FILTER FOR SELECTED PATIENTS WITH THYROID TEST AT TIME OF SURGERY
#--------------

tyroid_at_surgery_patients <- c(2,4,6,7,8,11,13,14,15,16)

names(meta) = gsub("[ /]", "_", names(meta))
serum_cols = grep("Serum", names(meta), value = T)
plot.meta = meta %>% select("Study.ID#", serum_cols)
plot.meta$Serum_Free_T3 = as.numeric(plot.meta$Serum_Free_T3)

plot.meta <- plot.meta %>% 
  # filter(!is.na(Serum_Free_T4), !is.na(Serum_TSH), !is.na(Serum_Free_T3)) %>%
  filter(`Study.ID#` %in% tyroid_at_surgery_patients)

all(tyroid_at_surgery_patients %in% plot.meta$`Study.ID#`)

plot.meta

# Plot with correlation labels in the top-left corner
for(col in (2:4)){
  col2 = 2
  while(col2 <= 4){
    var1 = names(plot.meta)[col]
    var2 = names(plot.meta)[col2]
    
    
    
    sp <- ggscatter(plot.meta, x = var1, y = var2,
      color = "black",
      add = "reg.line", 
      conf.int = T) #Change this to TRUE if you want to add the Coef interval
    
    max.data = plot.meta %>% select(all_of(c(var1, var2))) %>% filter(!is.na(!!sym(var1)), !is.na(!!sym(var2)))
    max.y = max(max.data[[var2]], na.rm = T)
    max.x = max(max.data[[var1]], na.rm = T)
    sp = sp + stat_cor(
      method = "spearman", cor.coef.name = "rho", 
      label.x = -Inf, label.y = c(max.y*1.1*1.05, max.y*1.1), hjust = -0.1, vjust = 0, size = 6,
      show.legend = F) + 
      
      ##The code below should be useless based on the theme_set above
      theme_classic() +
      theme(
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 18),
        plot.title = element_text(size = 24, face = "bold"),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 16)

      ) +
      labs(title = paste(correct_mg_string(var1), correct_mg_string(var2), sep = " x "),
           x = correct_mg_string(var1),
           y = correct_mg_string(var2)
           ) +
      scale_y_continuous(limits = c(0, max.y*1.2)) +
      scale_x_continuous(limits = c(0, max.x*1.05))
    


    print(sp)
    # ggsave(create.dir = T, 
    #        filename = file.path("./figures/Paper_figures/Figure_4/Tyroid_at_surgery_filter/correlation_plots", var1, paste0(var1, "_x_", var2, ".png")),
    #       device = "png", 
    #       plot = sp,
    #       width = 6, height = 6,
    #       dpi = 600)
    col2 = col2 + 1
  }
}

names(plot.meta)
grep("u", names(plot.meta))
```




